Principal Component Analysis of Population Genetic Data

### 2019-09-26 paper-bedpca.tex
save

When there is some missing values, we effectively compute the partial SVD of the following matrix: $\breve{G}_{i,j} = \frac{G_{i,j} - 2 f_j}{\sqrt{2 f_j (1 - f_j) ~ p_i^{(1)} p_j^{(2)}}}$, where missing values are replaced by the mean (i.e.\ $G_{i,j} - 2 f_j = 0$ if $G_{i,j}$ is missing) and where the proportion of missing values of individual $i$, $p_i^{(1)}$, and variant $j$, $p_j^{(2)}$, are accounting for the shrinkage due to mean imputation [REWORD?].
Note that this decomposition is equivalent to the decomposition presented above when there is no missing value.

\subsection{Projecting PCs of the 1000 Genomes}

@@

There are [X] steps to project PCs of a reference dataset (e.g.\ 1000G) to a target genotype data:
\begin{enumerate}
\item matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles;
\item computing PCA of the reference dataset using the matched variants only;
\item projecting computed PCs to the target data;
\item possibly adjusting projected PCs for shrinkage;
\item[5.] [ANYTHING ELSE?].
\end{enumerate}
We implement these steps in function \texttt{bed\_projectPCA} of package bigsnpr.

For step 4, we implement the shrinkage-bias adjustment proposed by \cite{dey2019asymptotic}.
This requires computing all eigenvalues of the reference dataset, which is relatively fast to compute for the 2493 individuals of the 1000G. However, it can be computionally expensive if the reference data used includes many more individuals [BETWEEN O($N^2$) AND O($N^3$); MENTION IT?].

@@

>>>>

When there is some missing values, we effectively compute the partial SVD of the following matrix: $\breve{G}_{i,j} = \frac{G_{i,j} - 2 f_j}{\sqrt{2 f_j (1 - f_j) ~ p_i^{(1)} p_j^{(2)}}}$ [VERIFY THAT THIS IS CENTERED], where missing values are replaced by the mean (i.e.\ $G_{i,j} - 2 f_j = 0$ if $G_{i,j}$ is missing) and where the proportion of missing values of individual $i$, $p_i^{(1)}$, and variant $j$, $p_j^{(2)}$, are accounting for the shrinkage due to mean imputation [REWORD?].
Note that this decomposition is equivalent to the decomposition presented above when there is no missing value.

\subsection{Outlier detection in PCA}

We propose [BOF] different outlier detection techniques for PCA.

For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use the same procedure as described by \cite{prive2017efficient}: we compute pcadapt statistics \cite[]{luu2017pcadapt} that summarises the contribution of each variant in all PC loadings. This captures both population-differentiating variants as well as long-range LD regions. We then smooth the statistics to mostly capture consecutive outliers that corresponds to long-range LD regions.

For detecting outlier samples in PCA, we use the Local Outlier Factor (LOF) on PCs \cite[]{breunig2000lof}. To make it robust to the choice of the number of nearest neighbours (kNN) and to the number of PCs used, we take the maximum value of the statistic computed with kNN $\in$ \{4, 10, 30\} and varying the number of PCs from 2 to the number of available PCs. We make use of the fast K nearest neighbours implementation of R package nabor to implement the LOF statistic efficiently \cite[]{elseberg2012comparison}.

For detecting samples that have a different ancestry from most of the samples in the data, we compute the pairwise orthogonalized Gnanadesikan-Kettenrin robust Mahalanobis distance of PCs \cite[]{gnanadesikan1972robust,maronna2002robust}, and do not include individuals whose log-distance is larger than some threshold determined based on visual inspection [DO IT AUTOMATIC?].

Finally, to choose the threshold of outlierness based on the previously described statistics, we use a modified version of Tukey's rule, a standard rule for detecting outliers \cite[]{tukey77}. 
The standard upper limit defined by Tukey's rule is $q_{75\%}(x) + 1.5 \cdot IQR(x)$, where $x$ is the vector of computed statistics and $IQR(x) = q_{75\%}(x) - q_{25\%}(x)$ is the interquartile range.
However, there are two pitfalls when using Tukey's rule. First, Tukey's rule assumes a normally distributed sample. For example, when the data is skewed, it does not work that well. We account for skewness using the medcouple \cite[]{hubert2008adjusted}. Standard Tukey's rule also uses a fixed coefficient (1.5) that does not account for multiple testing, which means that for large samples, you will almost always get some outliers if using 1.5. 
To solve these two issues, we implement \texttt{tukey\_mc\_up} in R package bigutilsr and use it here, which accounts both for skewness and multiple testing by default.

\subsection{Projecting PCs from a reference dataset}

@@

There are 4 steps to project PCs of a reference dataset (e.g.\ 1000G) to a target genotype data:
\begin{enumerate}
\item matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles;
\item computing PCA of the reference dataset using the matched variants only;
\item projecting computed PCs to the target data;
\item adjusting projected PCs for shrinkage.
\end{enumerate}
We implement these steps in function \texttt{bed\_projectPCA} of package bigsnpr.

For step 4, we implement a shrinkage-bias adjustment method that combines the two methods proposed by \cite{zhang2019fast}. First, for only a subset of target individuals, we compute the Online Augmentation, Decomposition, and Procrustes (OADP) transformation (using $K''=K'=K$). For these individuals, we regress the new PC scores resulting from OADP by the projected PC scores without adjustement; this gives us shrinkage-adjustment factors that we use to correct projected PC scores for all target individuals.

%This requires computing all eigenvalues of the reference dataset, which is relatively fast to compute for the 2493 individuals of the 1000G. However, it can be computionally expensive if the reference data used includes many more individuals [BETWEEN O($N^2$) AND O($N^3$); MENTION IT?].

@@

[LIMITATION: RUN MULTIPLE TIMES / FAIL TO SPEED UP]
